---
title: "Spatial Econometrics Models with Jags and Stan"
author: "Virgilio Gomez-Rubio"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Spatial Econometrics Models with Jags and Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The `SEMCMC` package implements a number of spatial econometrics models (LeSage and Pace, 2009)) that can be fitted using MCMC with `rjags` or `rstan`.

# Spatial Econometrics Models

Spatial econometrics models are a family of spatially autocorrelated models
where spatial autocorrelation can be on the responde, the error term or both.
The study region is divied into $n$ non-overlapping regions and the aim is to
model a vecotr of responses $y = (y_1,\ldots,y_n)$ using $k$ covariates. The
matrix of covariates  $X$ is then a matrix of dimension $n \times k$. Also,
adjacency is defined by an adjacency matrix $W$, which will row-standardised.

The first model that we will introduce is the *spatial error model* (SEM),
which is an autocorrelated model on the response:

$$
y = X \beta + u;\ u = \rho W u + e;\ e \sim N(0, \sigma^2 I)
$$

Note that the error term $u$ is spatially autocorrelated with spatial
autocorrelation parameter being $\lambda$.

This model can be re-written as

$$
y = X \beta + u;\ u \sim N\left(0, \sigma^2[(I -\lambda W^{\intercal})(I-\lambda W)]^{-1}\right)
$$


Similarly, the *spatial lag model* (SLM) is a spatially autocorrelated model
on the response:

$$
y = \rho W y + X \beta + e; e \sim N(0, \sigma^2 I)
$$

This model can be also written as

$$
y = (I - \rho W) ^{-1} X\beta + u; u \sim N\left(0, \sigma^2[(I -\rho W^{\intercal})(I-\rho W)]^{-1}\right)
$$

The *spatial Durbin model* (SDM) is a SLM model with lagged-covariates $WX$:

$$
y = (I - \rho W) ^{-1} (X\beta + W X \gamma)+ u; u \sim N\left(0, \sigma^2[(I -\rho W^{\intercal})(I-\rho W)]^{-1}\right)
$$


Finally the *general spatial model* (SAC) is a model that includes
autocorrelated terms on the error and the reponse:

$$
y = \rho W y + X \beta + \ u = \rho W u + e;\ e \sim N(0, \sigma^2 I)
$$

The SAC model can be rewritten as 

$$
y = (I - \rho W) ^{-1} X\beta + u; u \sim N\left(0, \sigma^2[(I -\rho W^{\intercal}) (I -\lambda W^{\intercal})(I-\lambda W) (I-\rho W)]^{-1}\right)
$$


In addition, the SAC model can include lagged covariates, so that it becomes:

$$
y = (I - \rho W) ^{-1} (X\beta + W X \gamma) + u; u \sim N\left(0, \sigma^2[(I -\rho W^{\intercal}) (I -\lambda W^{\intercal})(I-\lambda W) (I-\rho W)]^{-1}\right)
$$


The SLX model is essentially a SEM model with lagged covariates:

$$
y = X\beta + W X \gamma + u;\ 
u \sim N\left(0, \sigma^2[(I -\lambda W^{\intercal})(I-\lambda W)]^{-1}\right)
$$


In addition to these spatial econometrics models, `SEMCMC`also implements
the proper CAR model on the error term:

$$
y = X\beta + u; u \sim N\left(0, \sigma^2(I-\lambda W)^{-1}\right)
$$

# Bayesian inference with the  `SEMCMC` package


The `SEMCMC` includes an implementation of these models using Markov chain
Monte Carlo (MCMC) methods with jags and Stan. So far, available models
are SEM, SLM, SDM, SAC, SAC with lagged covariates and SLX for a Gaussian
response.

Models are fitted with a call to `SEMCMC()` and it allows the following:

* Covariates and response variables are specified with a `formula`.

* The model to fit is set using argument `model`.

* The sampler (either `jags` or `stan`) can be set with argument `sampler`.

* The number of iterations for burn-in and inference can also be set.

* Computation of the impacts is available via the `impacts` function
(currently, only available for jags output).

Non-linear models will be implemented soon and some are already available for
jags.

The priors used for the parameters are as follows:

* Covariate coefficients are assigned a Gaussian prior with zero mean and precision 0.001.

* Precisions are assigned a Gamma distribution with parameters 0.01 and 0.01.

* Spatial autocorrelation parameters are assigned a uniform prior between
the inverse of the minimum and the inverse of the maximum eigenvalues of
$W$.

Other priors could be easily used by modifying the model definitions 
(available under `SEMCMC/inst`).

# Example: Columbus (OH) dataset

In this example we will use the `columbus` dataset to  illustrate the use
of the `SEMCMC` function to fit Bayesian spatial econometrics models.
This dataset is available in the `spdep` package and it contains
informatin about 49 neighbourhoods in Columbus, OH, in 1980.

In particular, we will be using the following three variables:

* `HOVAL` housing value (in \$1,000)

* `INC` household income (in \$1,000)

* `CRIME` residential burglaries and vehicle thefts per thousand
          households in the neighborhood

The aim is to model crime rates on housing value and crime using different models.


First of all, data needs to be loaded and the adjacency matrix converted into
a matrix $W$:

```{r}
library(spdep)
data(columbus)
     
W <- nb2mat(col.gal.nb, style = "W")
```

Next, the response variable and covariates are set using a formula:

```{r}
m.form <-  CRIME ~ INC + HOVAL
```

This formula will be used in all the models that we fit below.


```{r echo = FALSE, results = "hide", message = FALSE}
library(SEMCMC)
#Load precomputed results
load("SEMCMC.RData")
```

In the follwoing examples, we willbe fittind models SEM, SDM and SAC with lagged
covariates. Other models can be fitted in a similar way by setting argument
`model` to the appropriate value. See the manual page of `SEMCMC` for some examples.

## SEM model

We will start with the SEM model, using both jags and stan:

```{r results = "hide", eval = FALSE}
library(SEMCMC)
#JAGS
sem.jags <- SEMCMC(m.form, data = columbus, W = W, model = "sem", sampler = "jags", n.burnin = 2000, n.iter = 200, n.thin = 25)
#STAN
sem.stan <- SEMCMC(m.form, data = columbus, W = W, model = "sem", sampler = "stan", n.burnin = 2000, n.iter = 200, n.thin = 25)
```

By default, the number of burn-in iterations is 1000 and the number of
iterations for inference is 1000 with no thinning.

Note how the only difference is in `sampler`. In addition, we will use 
maximum likelihood to fit this model so that we can compare the different
estimates:

```{r}
#Maximum likelihood
#Adjacacy matrix as a list of weights (see 'listw')
lw <- nb2listw(col.gal.nb, style="W")
sem.ml <- errorsarlm(m.form, data = columbus, lw, method="eigen", 
  quiet = TRUE)
```


```{r echo = FALSE}
#Some useful functions...

#stan2coda, by Ben Goodrich
#http://jeromyanglim.tumblr.com/post/91434443911/how-to-convert-a-stan-fit-object-to-work-with-coda
stan2coda <- function(fit) {
     mcmc.list(lapply(1:ncol(fit), function(x) mcmc(as.array(fit)[,x,])))
}


#Summary output of jags output
summary_output <- function (fit, func) {
  apply(SEMCMC:::jags2coda(fit)[[1]], 2, mean)
}

```

The point estimates of the parameters for the SEM model are available in the
following table:

```{r echo = FALSE}
#SEM summary output

tab.sem <- data.frame(JAGS = summary_output(sem.jags$results, mean))
tab.sem$STAN <- summary(stan2coda(sem.stan$results))$stat[1:5, 1]
tab.sem$ML <- c(sem.ml$coefficients, sem.ml$lambda, 1/sem.ml$s2)

#Update covariate names
var.names <- names(sem.ml$coefficients)
row.names(tab.sem)[1:length(var.names)] <- var.names 

knitr::kable(tab.sem)
```


## SDM model

```{r results = "hide", eval = FALSE}
library(SEMCMC)
#JAGS
sdm.jags <- SEMCMC(m.form, data = columbus, W = W, model = "sdm", sampler = "jags", n.burnin = 2000, n.iter = 200, n.thin = 25)
#STAN
sdm.stan <- SEMCMC(m.form, data = columbus, W = W, model = "sdm", sampler = "stan", n.burnin = 2000, n.iter = 200, n.thin = 25)
#ML
```

```{r}
sdm.ml <- lagsarlm(m.form, data = columbus, lw, type = "mixed", 
  method = "eigen", quiet = TRUE)
```


For the spatial Durbin model, the point estimates are the following:

```{r echo = FALSE}
#SDM summary output
tab.sdm <- data.frame(JAGS = summary_output(sdm.jags$results, mean))
tab.sdm$STAN <- summary(stan2coda(sdm.stan$results))$stat[1:7, 1]
tab.sdm$ML <- c(sdm.ml$coefficients, sdm.ml$rho, 1/sdm.ml$s2)

#Update covariate names
var.names <- names(sdm.ml$coefficients)
row.names(tab.sdm)[1:length(var.names)] <- var.names


knitr::kable(tab.sdm)
```



##SAC model with lagged covariates

```{r results = "hide", eval = FALSE}
#JAGS
sacmixed.jags <- SEMCMC(m.form, data = columbus, W = W, model = "sacmixed",
  sampler = "jags", n.burnin = 2000, n.iter = 200, n.thin = 25)
#STAN
sacmixed.stan <- SEMCMC(m.form, data = columbus, W = W, model = "sacmixed",
  sampler = "stan", n.burnin = 2000, n.iter = 200, n.thin = 25)
```

```{r}
#ML
sacmixed.ml <- sacsarlm(m.form, data = columbus, lw, type = "sacmixed", 
  method = "eigen", quiet = TRUE)
```


The point estimats of the parameters in the model for the SAC model with lagged
covariates are:

```{r echo = FALSE}
#SAC mixed summary output
tab.sacmixed <- data.frame(JAGS = summary_output(sacmixed.jags$results, mean))
tab.sacmixed$STAN <- summary(stan2coda(sacmixed.stan$results))$stat[1:8, 1]
tab.sacmixed$ML <- c(sacmixed.ml$coefficients, sacmixed.ml$lambda, sacmixed.ml$rho, 1/sacmixed.ml$s2)

#Update covariate names
var.names <- names(sacmixed.ml$coefficients)
row.names(tab.sacmixed)[1:length(var.names)] <- var.names


knitr::kable(tab.sacmixed)
```

# Impacts


Computation of the impacts is available for all implemented models via the
`impacts()` function. This will provide point estimates (i.e., posterior
means) and posterior standard deviations of the impacts.

## SEM model

```{r}
impacts(sem.jags)
impacts(sem.stan)
```

## SDM model


```{r}
impacts(sdm.jags)
impacts(sdm.stan)
spdep::impacts(sdm.ml, listw = lw)
```

## SAC model with lagged covariates


```{r}
impacts(sacmixed.jags)
impacts(sacmixed.stan)
spdep::impacts(sacmixed.ml, listw = lw)
```

